D3064R0: How to Avoid OOTA Without Really Trying

### 2024-02-20 cuttingroom.tex
Apply Feb 18 Alan Stern feedback, part 2
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

>>>>

In happy contrast to the situation at the start of the C++11 memory-model
effort, heavily used CPU families now have accurate and executable formal
memory models, all of which prohibit the causal cycles required to form
OOTA cycles.
Even more important, there is clarity on exactly what hardware can and
can not do in the course of speculative execution.
Finally, computation is based on instructions, and instructions require
finite time to execute, even when executed speculatively.

### 2024-02-20 ootavacuous.tex
Apply Feb 18 Alan Stern feedback, part 1
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Apply Feb 18 Alan Stern feedback, part 2
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

This paper shows that C++ implementations subject to reasonable
real-world constraints cannot exhibit OOTA cycles.
Section~\ref{sec:Background}
provides background on the problem of OOTA cycles, including a
brief summary of prior work.
Section~\ref{sec:OOTA and Semantic Dependencies}
motivates a shift in definition of OOTA cycle and reviews properties of
semantic dependencies, showing that these properties are a function of
executions rather than simply of the source code.
This leads into
Section~\ref{sec:What is an Execution?},
which describes properties of abstract and hardware executions, along
with the relationship between the two.

@@

In happy contrast to the situation at the start of the C++11 memory-model
effort, heavily used CPU families now have accurate and executable formal
memory models, all of which prohibit the causal cycles required to form
OOTA cycles.
Even more important, there is clarity on exactly what hardware can and
can not do in the course of speculative execution.
Finally, computation is based on instructions, and instructions require
finite time to execute, even when executed speculatively.

@@

For example, in Listing~\ref{lst:Speculated Store and Non-Speculated Load},
suppose that the storage for \co{x} is located immediately before that
for \co{y[]}.
Suppose further that hardware speculation incorrectly guesses the value
of \co{i} to be -1, so that the speculated store of 42 on line~9 uses
the address of \co{x}.\footnote{
	Yes, \co{y[-1]} is UB, but the CPU neither knows
	nor cares, nor should it.}
Then line~11 can cause the value 42 to be loaded into \co{r1}, when
by the rules of the C++ abstract machine the value must instead be
either zero or one.

@@

Note that if both legs of the \co{if} statement spanning lines~9-12 were
to store the same value, the implementation would be within its rights
to hoist the store out of that \co{if} statement.
However, in this case, the two stores on lines~10 and~12 are storing
distinct values, preventing the hoisting.

Note also that some architectures provide conditional-move instructions,
which allows line~9 to avoid emitting a conditional branch.
However, such architectures would still treat a store of the
target of the conditional-move instructions as speculative until
such time as the conditional moves (and any loads that they depend
on) complete.

@@

The key point is that a store having a semantic dependency on a given
set of loads must execute after the completion of the last of those
loads in global time.

>>>>

This paper shows that compiler-based C++ implementations subject to
reasonable real-world constraints cannot exhibit OOTA cycles.
Section~\ref{sec:Background}
provides background on the problem of OOTA cycles, including a
brief summary of prior work.
Section~\ref{sec:OOTA and Semantic Dependencies}
motivates a shift in definition of OOTA cycle and reviews properties of
semantic dependencies, showing that these properties are a function of
executions rather than simply of the source code.
This leads into
Section~\ref{sec:What is an Execution?},
which describes properties of abstract and hardware executions, along
with the relationship between the two.

@@

On real systems, instructions require time to execute, even
when executed speculatively.
For example, in Listing~\ref{lst:Speculated Store and Non-Speculated Load},
suppose that the storage for \co{x} is located immediately before that
for \co{y[]}.
Suppose further that hardware speculation incorrectly guesses the value
of \co{i} to be -1, so that the speculated store of 42 on line~9 uses
the address of \co{x}.\footnote{
	Yes, \co{y[-1]} is UB, but the CPU neither knows
	nor cares, nor should it.}
Then line~11 can cause the value 42 to be loaded into \co{r1}, when
by the rules of the C++ abstract machine the value must instead be
either zero or one.

@@

@@

One could imagine a conditional-store instruction, similar to existing
conditional-move instructions, but with condition controlling a store
to memory rather than a move to a register.
The condition would control whether or not the store was executed,
and prior conditional-move instructions could control the value
stored and the address stored to.
With or without prior conditional-move instructions, but assuming a
compiler doing single-thread analysis, a conditional-store instruction
permits generation of code corresponding to lines~10 and~11 of
Listing~\ref{lst:Causality Test Case 2}
on page~\pageref{lst:Causality Test Case 2}
without use of a conditional branch.
This means that the control dependency would extend from the loads on
lines~8 and~9 to the store on line~11 and nowhere else, potentially
enabling more hardware optimizations than would be permitted for code
relying on conditional branches.
Nevertheless, the conditional store could not be committed while any
the instructions it depends on were still being speculatively executed.

@@

The key point is that a store having a semantic dependency on a given
set of loads must execute after the completion of at least some of those
loads in global time.
